---
title: "Github Open Data Api Demo"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document contains embedded R code that demonstrates how to access FI$Cal expenditure data located at the [California Open Data Portal](https://data.ca.gov/) using the CKAN API.

The .Rmd version of this document can be knit within RStudio to produce the .md version of this document (both included in this repo).

See also the documentation for the [CKAN API](https://docs.ckan.org/en/ckan-2.7.3/api/) and the [CKAN Datastore API](https://docs.ckan.org/en/ckan-2.7.3/maintaining/datastore.html#the-datastore-api).

## Load Necessary Packages

If you don't have these packages installed, install them with `install.packages("packagename")`.
```{r load, results = "hide", message = F, warning = F}
library(jsonlite)  #JSON parser and generator
library(tidyverse) #a set of packages that make working in R more efficient
library(data.table) #more efficient extension of R's base data.frames
```

## Get Resource IDs

This script uses the CKAN API to get all the information about the vendor-transactions package (i.e. dataset) and extract the IDs for each of the resources that can be searched with the datastore API. For this particular dataset, there is one resource (i.e. file) for each month of data.

This is an important step, as the resulting list of resource IDs will allow future scripts to loop through all of the resources to get information on all months.

Note there are two FI\$Cal datasets available on the Open Data Portal:

* [Vendor Transactions](https://data.ca.gov/dataset/vendor-transactions) is the smaller dataset, containing all transactions for which there is an associated vendor name.
* [Spending Transactions](https://data.ca.gov/dataset/spending-transactions) is the larger dataset, containing every expenditure transaction for the more than 150 departments that use FI\$Cal for their accounting.

The code examples in this file use the Vendor Transactions dataset. To use the Spending Transactions dataset, replace the URL in the following code block with the URL in the commented line. All field names are identical between the datasets, except that Spending Transactions does not contain a "Vendor Name" field.

```{r get_resources}

### Link to the api 
url <- "https://data.ca.gov/api/3/action/package_show?id=vendor-transactions" #specifies that the package we want information for is "vendor-transactions"

# For "spending-transactions", replace the URL with: "https://data.ca.gov/api/3/action/package_show?id=spending-transactions", then follow the same steps below

### Use fromJSON to read the text in the link
vendor <- fromJSON(url)
vendor_resources <- vendor$result$resources #extracts just the results$resources table from the response
vendor_resource_ids <- vendor_resources$resource_id[which(vendor_resources$datastore_active)] #creates a vector of just the resource IDs that have indexed, searchable data (are "datastore active") - WILL BE USED LATER
vendor_resource_ids
```

## Code Demos Using One Resource

### Get the first 5 rows of data, from a specific resource.

```{r five_rows}
#Use resource ID for FY19P01
five_rows_url <- "https://data.ca.gov/api/3/action/datastore_search?resource_id=23460e28-6a4b-419d-997e-7c51251e733e&limit=5"
five_rows_JSON <- fromJSON(five_rows_url)
five_rows_records <- five_rows_JSON$result$records
five_rows_records
```

### Get all records containing `jones` from a specific resource.

```{r jones}
jones <- "https://data.ca.gov/api/3/action/datastore_search?resource_id=23460e28-6a4b-419d-997e-7c51251e733e&q=jones"
jones_JSON <- fromJSON(jones)
jones_records <- jones_JSON$result$records
head(jones_records)
```

### Use SQL to get all account codes like `5442000` from a specific resource.

Account code 5442000 is Medical and Health Care Payments.

```{r test_account}
test_account <- "https://data.ca.gov/api/3/action/datastore_search_sql?sql=SELECT%20*%20from%20%2223460e28-6a4b-419d-997e-7c51251e733e%22%20WHERE%20%22account%22%20LIKE%20%275442000%27"
test_account_JSON <- fromJSON(test_account)
test_account_records <- test_account_JSON$result$records
head(test_account_records)
```


## Use a loop to get data from all resources

Using the `vendor_resource_ids` vector we created earlier, this script loops through all of the resource IDs and gets information for the specified account code for all months. This example uses Account Code 5326100, which is "electricity".

Note that by default, a query will return no more than 100 rows. However, the limit can be adjusted to return up to 32,000 rows per query. This code adjusts the limit to the maximum and then checks whether the total number of records that match the query is greater than 32,000. If it is, the code loops through the query as many times as necessary to return all the matching records from a given resource before moving on to the next.

```{r loop_through_resources}
FISCal_list <- vector(mode = "list", length = length(vendor_resource_ids)) #Creates a list to drop the retrieved data into; in this case, it will be a list of data frames

account_code_query <- "&q=5326100" #This is the account code for "Electricity"

count = 0

# Use a for loop to iterate through each resource in the list
for (item in vendor_resource_ids[37:length(vendor_resource_ids)]) {
  count = count + 1
  templist <- try(fromJSON(paste0("https://data.ca.gov/api/3/action/datastore_search?&resource_id=", item, "&limit=32000", account_code_query)), silent = T)
  tempDF <- templist$result$records # Creates a temporary data frame to store results
  row_total <- templist$result$total # Stores the number of records in the resource that match the query, which might be more than the 32K row max query limit
  resource_loops = 1
  #The following loop only runs if the number of matching records exceeds the 32K row max query limit, adding records until all the records that match the query are in the data frame
  while (row_total > nrow(tempDF)) {
    offset = 32000 * resource_loops
    templist <- try(fromJSON(paste0("https://data.ca.gov/api/3/action/datastore_search?&resource_id=", item, "&limit=32000", account_code_query, "&offset=", offset)), silent = T)
    tempDF <- rbind(tempDF, templist$result$records)
    resource_loops = resource_loops + 1
  }
  FISCal_list[[count]] <- tempDF #this drops the resulting data frame into the appropriate spot in the list
  
}
FISCal <- rbindlist(FISCal_list) #data.table method to quickly combine all list items into one data frame
```


## Group and graph data from previous query

```{r group_and_graph}
options(scipen=999) #prevent scientific notation

electricity_spend <- FISCal[, c(10, 14, 16)] #extract date, agency, and monetary amount columns
electricity_spend$accounting_date <- as.Date(electricity_spend$accounting_date) #format accounting_date as date
electricity_spend$monetary_amount <- as.numeric(electricity_spend$monetary_amount) #format monetary_amount as numeric

### Summarizes dollar amount by month and agency
electricity_spend_sum <- electricity_spend %>%
  mutate(accounting_date = format(accounting_date, "%Y-%m")) %>%
  group_by(accounting_date, agency_name) %>%
  summarize(monthly_total = sum(monetary_amount), .groups = 'drop')

### Stacked bar plot by monthly monetary amount for different agencies
ggplot(data=electricity_spend_sum,
       aes(fill=agency_name, x=accounting_date, y=monthly_total))+
  geom_bar(position="stack", stat="identity")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  ggtitle("Electricity Spending By State Agency Over Time")+
  xlab("Accounting Date") + ylab("Monthly Expenditure ($)")
```


